---
title: "Mesa City Traffic Arrest Violation Prediction"
author: "Shashank Sanjee Venkata Chalapathi"
date: "11/08/2020"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Required Packages 

```{r ,warning=FALSE,message=FALSE}

library(readr)
library(plotly)
library(tidyverse)
library(lubridate)
library(MASS)
library(mice)
library(ROSE)
library(caret)
library(randomForest)
library(e1071)
library(partykit)
library(nnet)
library(revgeo)
library(opencage)
library(rgeos)
library(ggforce)
library(ggplot2)
library(ggmap)
library(DT)
library(viridis)
library(rpart.plot)
library(patchwork)

```

### Reading data and cleaning

```{r,warning=FALSE,message=FALSE}

az_me <- read_csv("az_mesa_2019_12_17.csv")
population<-read_csv("Population.csv")
zipcode<-read_csv("zipcode.csv")
az_me <- az_me[,c("date","time","subject_age","subject_race","subject_sex","violation","arrest_made","lng","lat","type")]
az_me<- az_me %>% filter( type=="vehicular" & subject_race !="unknown" )

az_me<-na.omit(az_me)
```

## Preliminary Data Analysis


### Trend for stopping accross mesa 

```{r ,warning=FALSE,message=FALSE}
graph<-az_me %>%
  group_by(Year=year(date),subject_race)%>%count() 
graph <- inner_join(graph,population,by=c('Year','subject_race')) 

graph<-graph %>% mutate(percentage=n*100/Total)

g1<-ggplot(graph, aes(
  x = Year,
  y = n,
  col= subject_race
  )) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("Total Number of cases") +
  theme(axis.text.x = element_text(face = "bold", size = 10, angle = 45))
g2<-ggplot(graph, aes(
  x = Year,
  y = percentage,
  col= subject_race
  )) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("Percentage of population by race") +
  theme(axis.text.x = element_text(face = "bold", size = 10, angle = 45))

combined <- g1 + g2 & theme(legend.position = "right") 
combined + plot_layout(guides = "collect")

```

### Trend for arresting accross mesa

```{r,warning=FALSE,message=FALSE}
graph<-az_me %>% filter(arrest_made==TRUE) %>%
  group_by(Year=year(date),subject_race)%>%count() 
graph <- inner_join(graph,population,by=c('Year','subject_race')) 

graph<-graph %>% mutate(percentage=n*100/Total)

g1<-ggplot(graph, aes(
  x = Year,
  y = n,
  col= subject_race
  )) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("Total Number of cases") +
  theme(axis.text.x = element_text(face = "bold", size = 10, angle = 45))
g2<-ggplot(graph, aes(
  x = Year,
  y = percentage,
  col= subject_race
  )) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("Percentage of population by race") +
  theme(axis.text.x = element_text(face = "bold", size = 10, angle = 45))

combined <- g1 + g2 & theme(legend.position = "right") 
combined + plot_layout(guides = "collect")
```


### Total number of cases: Hour-wise

```{r ,warning=FALSE,message=FALSE}
az_me_age <- az_me %>%
mutate(age_level=cut(az_me$subject_age, breaks =c(15,30,50,81), labels=c('15-30','30-50','>50')))
az_me_age$subject_race=ifelse(az_me_age$subject_race=='unknown',NA,az_me_age$subject_race)
az_me_age<-na.omit(az_me_age)

graph_time<-az_me_age %>%
  group_by(hour=as.factor(hour(time)),age_level)%>%count()

g2<-ggplot(na.omit(graph_time), aes(
  x = hour,
  y = n,
  fill= age_level
  )) +
  geom_col() +
  xlab("Hour") +
  ylab("Count") +
  theme(axis.text.x = element_text(face = "bold", size = 10, angle = 45))+
  scale_fill_brewer(palette = "Dark2") 
ggplotly(g2)
```

### Total number of cases: Day - Wise

```{r ,warning=FALSE,message=FALSE}
graph_day<-az_me_age %>%
  group_by(day=as.factor(weekdays(as.Date(date))),age_level)%>%count()
# graph_day$subject_race <- fct_relevel(as.factor(graph_day$subject_race),'white',after=0L)

g3<-ggplot(graph_day, aes(
  x = reorder(day,n),
  y = n,
  fill= age_level
  )) +
  geom_col(width = 0.5) +
  xlab("Days") +
  ylab("Count")+
  theme(axis.text.x = element_text(face = "bold", size = 10, angle = 45))+
  scale_fill_brewer(palette = "Dark2") 
ggplotly(g3)
```

### Total number of vehicles stopped vs Reason for being stopped

```{r,,warning=FALSE,message=FALSE}


violation<-az_me%>%filter(type=="vehicular") %>% group_by(violation)%>%summarise(count=n())
violation <- na.omit(violation)
violation<-violation[order(violation$count,decreasing = TRUE),]

df3<-az_me_age %>%
  filter(type=="vehicular") %>%
      group_by(violation,age_level) %>%
      summarise(count=n())%>%
      arrange(desc(count))
    
    df3<-na.omit(df3)
    df3$violation<-as.factor(df3$violation)
    df4<-df3[1:15,]
    g4<-ggplot(data=df4, aes(x=reorder(violation,age_level),y=count,fill=age_level)) +
      geom_bar(stat = "identity",position = "dodge") +
      coord_flip()+ 
      xlab("Reason for being stopped") +
      ylab("Total number of vehicles stopped")
      
 ggplotly(g4)
```


## Model fitting using caret package



#### Selecting only arrest data 
```{r}

az_me_age_model<-az_me %>% filter( arrest_made == TRUE)

```


#### Imputing zipcode which is generated from latitude and longitude values which will be a predictor for my model
```{r,warning=FALSE,message=FALSE}
# Code for generating zipcode is commented because it gets stuck while knitting the file. so I have written the file to directory and read in to this rmd separately
# x<-revgeo(az_me_age_model$lng, az_me_age_model$lat,output = "frame")$zip
# y<-str_replace(x, ".*\\b(\\d{5})\\b.*", "\\1")
# write.csv(y,"~/Project/zipcode.csv", row.names = FALSE)
az_me_age_model$zipcode<-zipcode$x
az_me_age_model[az_me_age_model$zipcode=="Postcode Not Found",]$zipcode=NA
az_me_age_model<-na.omit(az_me_age_model)
```

### Dividing violations in 5 broad categories which act as response for my model
```{r,warning=FALSE,message=FALSE}

az_me_age_model$viotype <-
  ifelse(
    grepl("DRUGS|LIQUOR|PHONE", az_me_age_model$violation),
    "distracted_driving",
       ifelse(
        grepl("DEVICE|LIGHT|STOP|LIGHTS|LAMPS", az_me_age_model$violation),
        "running_red_light_or_without_headlight",
        ifelse(
          grepl(
            "TURN|SPEED|MPH|FOLLOWING|AGGRESSIVE |POLICE|DRIVEWAY|JAYWALKING|PEDESTRIAN|RECKLESS|SIDEWALK|PSS|LEFT|RIGHT|SIGNAL|LANE|UNSAFE|EMERGENCY",
            az_me_age_model$violation
          ),
          "reckless_driving",
          ifelse(
            grepl(
              "ACCIDENT|QUICK|UNATT|INFO|REPORT|DAMAGE|RELEASE",
              az_me_age_model$violation
            ),
            "leaving_scene_of_an_accident_informing",
            "without_proper_documents_or_vehicle_defects"
          )
        )
      )
    )
  
  
az_me_age_model<-az_me_age_model %>%
  mutate(subject_race=as.factor(subject_race),
         subject_sex=as.factor(subject_sex),
         viotype=as.factor(viotype),
         zipcode=as.factor(zipcode))

```

## Types of Models fitted

### Multinomial Model 

```{r,warning=FALSE,message=FALSE,results="hide"}
model<- viotype~subject_age+subject_race+subject_sex+zipcode

#Multinomial model with up sampling
ctrl <- trainControl(method = "repeatedcv", 
                     number = 10,
                     sampling = "up")

set.seed(42)
model_rf_over <- caret::train(model,
                              data = az_me_age_model,
                              method = "multinom",
                              preProcess = c("scale", "center"),
                              trControl = ctrl,
                              tuneLength = 5)
```



```{r,,message=FALSE,warning=FALSE}
x<-predict(model_rf_over, newdata = az_me_age_model, type = "raw")
result<- confusionMatrix(x,az_me_age_model$viotype)
result

#For combinig the Recall Precision to calculate F1 score
get.macro.f1 <- function(cm) {
    c <- cm$byClass # a single matrix is sufficient
    re <- sum(c[, "Recall"]) / nrow(c)
    pr <- sum(c[, "Precision"]) / nrow(c)
    f1 <- 2 * ((re * pr) / (re + pr))
    ac<-cm$overall["Accuracy"]
    f2<-c(ac,re,pr,f1)
    return(f2)
}
macro.mul <- get.macro.f1(result)

```

### Decision tree 
```{r,warning=FALSE,message=FALSE}
ctrl <- trainControl(method = "repeatedcv", 
                     number = 10,
                     repeats = 3,
                     sampling = "up")

set.seed(42)
model_rf_over <- caret::train(model,
                              data = az_me_age_model,
                              method = "rpart",
                              parms = list(split = "information"),
                              tuneLength = 10,
                              trControl = ctrl)

x<-predict(model_rf_over, newdata = az_me_age_model, type = "raw")
result<-confusionMatrix(x,az_me_age_model$viotype)
result
get.macro.f1 <- function(cm) {
    c <- cm$byClass # a single matrix is sufficient
    re <- sum(c[, "Recall"]) / nrow(c)
    pr <- sum(c[, "Precision"]) / nrow(c)
    f1 <- 2 * ((re * pr) / (re + pr))
    ac<-cm$overall["Accuracy"]
    f2<-c(ac,re,pr,f1)
    return(f2)
}
macro.dt <- get.macro.f1(result)
```

### Random Forest 
```{r,warning=FALSE,message=FALSE}
ctrl <- trainControl(method = "oob", 
                     number = 10,
                     sampling = "up")

set.seed(42)
model_rf_over_rf <- caret::train(model,
                              data = az_me_age_model,
                              method = "rf",
                              trControl = ctrl)


x<-predict(model_rf_over_rf, newdata = az_me_age_model, type = "raw")
result<-confusionMatrix(x,az_me_age_model$viotype)
result

get.macro.f1 <- function(cm) {
    c <- cm$byClass # a single matrix is sufficient
    re <- sum(c[, "Recall"]) / nrow(c)
    pr <- sum(c[, "Precision"]) / nrow(c)
    f1 <- 2 * ((re * pr) / (re + pr))
    ac<-cm$overall["Accuracy"]
    f2<-c(ac,re,pr,f1)
    return(f2)
}
macro.rand <- get.macro.f1(result)
```


### Neural Net 
```{r,message=FALSE,warning=FALSE,results="hide"}
ctrl <- trainControl(method = "repeatedcv", 
                     number = 10, 
                     verboseIter = FALSE,
                     sampling = "up")

set.seed(42)
model_rf_over <- caret::train(model,
                              data = az_me_age_model,
                              method = "nnet",
                              maxit = 1000,
                              trControl = ctrl)

```

```{r,,message=FALSE,warning=FALSE}
x<-predict(model_rf_over, newdata = az_me_age_model, type = "raw")
result<-confusionMatrix(x,az_me_age_model$viotype)
result
get.macro.f1 <- function(cm) {
    c <- cm$byClass # a single matrix is sufficient
    re <- sum(c[, "Recall"]) / nrow(c)
    pr <- sum(c[, "Precision"]) / nrow(c)
    f1 <- 2 * ((re * pr) / (re + pr))
    ac<-cm$overall["Accuracy"]
    f2<-c(ac,re,pr,f1)
    return(f2)
}
macro.nn <- get.macro.f1(result)

```


### Knn  
```{r,message=FALSE,warning=FALSE}
ctrl <- trainControl(method = "repeatedcv", 
                     repeats = 3,
                     sampling = "up")

set.seed(42)
model_rf_over <- caret::train(model,
                              data = az_me_age_model,
                              method = "knn",
                              preProcess = c("scale", "center"),
                              tuneLength = 20,
                              trControl = ctrl)

x<-predict(model_rf_over, newdata = az_me_age_model, type = "raw")
result<-confusionMatrix(x,az_me_age_model$viotype)
result
get.macro.f1 <- function(cm) {
    c <- cm$byClass # a single matrix is sufficient
    re <- sum(c[, "Recall"]) / nrow(c)
    pr <- sum(c[, "Precision"]) / nrow(c)
    f1 <- 2 * ((re * pr) / (re + pr))
    ac<-cm$overall["Accuracy"]
    f2<-c(ac,re,pr,f1)
    return(f2)
}
macro.knn <- get.macro.f1(result)
rbind(macro.mul,macro.dt,macro.rand,macro.nn,macro.knn)

```


## Map visualizations

### Model Output with max probability violation 
```{r,warning=FALSE,message=FALSE}
#Importing map for mesa city
register_google(key = "AIzaSyCvgTagZzhtN_1FiBYEQy29kgkSYgxEAao", write = TRUE)
mesa <- ggmap(get_map(c(left=-111.9363,bottom=33.27729,right=-111.5822,top=33.49722)))

#Predicted probability percentages for getting arrested for different  violation types at different locations.
x<-predict(model_rf_over_rf, newdata = az_me_age_model, type = "prob")


#combining predicted values with original data set.
n<-nrow(x)
x$obs<-1:n
x_long <- x %>% pivot_longer(-obs,
                       names_to = "violation_type",
                       values_to = "prob")
x_max <- x_long %>%
            group_by(obs) %>%
              summarise(max_prob = max(prob),
                        violation_type = violation_type[which(prob == max_prob)])


az_me_age_model$obs<-c(1:n)
az_me_age_model<-inner_join(az_me_age_model,x_max,by="obs")
az_me_age_model_shiny<-inner_join(az_me_age_model,x_long,by="obs")

az_me_age_model <- az_me_age_model %>%
  arrange(obs) %>%
  mutate( violation_type=factor(violation_type, unique(violation_type))) %>%
  mutate( mytext=paste(
    "Violation Type: ", violation_type, "\n", 
    "Percentage of getting arrested for this violation: ", formatC(max_prob*100), sep="")
  )

shiny_data<- az_me_age_model_shiny %>% group_by(violation_type.y,subject_race,subject_sex) %>% distinct()
shiny_data$violation_type<- as.factor(shiny_data$violation_type.y)
shiny_data<-shiny_data[,c("subject_age","subject_race","subject_sex","lng","lat","violation_type","prob")]
shiny_data<-unique(shiny_data)
write.csv(shiny_data,"~/shiny_data.csv", row.names = FALSE)

# Make the map (static)
p <- mesa +
  geom_point(
    aes(
      x = lng,
      y = lat,
      text = mytext,
      color =violation_type
    ),
    data = az_me_age_model
  ) +
  theme_void() +
  ggtitle("Probability of arrest violation types across the MESA(AZ state)")+
  theme_bw() + theme(axis.title.x=element_blank(),
                     axis.title.y=element_blank(),
                     panel.border = element_blank())

p <- ggplotly(p,tooltip = "text")
p
```

### Heat map for number of arrests which are predicted from the model

```{r,message=FALSE,warning=FALSE}
az_me_age_model$lon<-az_me_age_model$lng
mesa +
  coord_equal() + 
  xlab('Longitude') + 
  ylab('Latitude') + 
  stat_density2d(aes(fill = ..level..), alpha = .3,
                 geom = "polygon", data = az_me_age_model,show.legend=TRUE) + 
  scale_fill_viridis_c() + 
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
        axis.text.y=element_blank(),axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank())+ labs(fill = "Number of Arrest Cases")+
  facet_wrap(~violation_type)

```
